Investigation of layer specific BOLD in the human visual cortex during visual attention

### 2021-01-31 paper.tex
updates before rxiv

\contrib[\authfn{1}]{These authors contributed equally to this work}
\contrib[\authfn{2}]{These authors also contributed equally to this work}

@@

\begin{abstract}
Directing spatial attention towards a particular stimulus enhances cortical responses at corresponding regions in cortex. How attention modulates the laminar response profile within the attended region, however, remains unclear. In this paper, we use high field (7T) fMRI to investigate the effects of attention on laminar activity profiles in areas V1-V3; both when a stimulus was presented to the observer, and in the absence of visual stimulation. Replicating previous findings, we find robust increases in the overall BOLD response for attended regions in cortex, both with and without visual stimulation. Interestingly, when analyzing the BOLD response across the individual layers in visual cortex, we observed no evidence for laminar-specific differentiation with attention. We offer several potential explanations for these results, including theoretical, methodological and technical reasons. Additionally, we provide all data and pipelines openly, in order to promote analytic consistency across layer-specific studies, improve reproducibility, and decrease the false positive rate as a result of analytical flexibility.
\end{abstract}

@@

It is known from anatomical studies that the human cerebral cortex can be subdivided into histological layers with different cell types. The cytoarchitectonic structure varies across the brain and forms the basis of the Brodmann atlas \citep{Brodmann1909}. While the precise function of each cortical layer remains unclear, their connectivity profile suggests a division in terms of bottom-up and top-down processing \citep{Felleman1991,Barone2000,Shipp2016}. Most brain areas have six different histological layers. Specifically, Layer IV and to a lesser extent Layer V/VI are commonly associated with receiving feedforward drive from Layer III of lower cortical areas or from the thalamus \citep{Jones1998,Constantinople2013}. Layers I-II and VI, in contrast, are typically implicated in receiving downward information flow (feedback), which often originates from layer V \citep{Alitto2003}. Interestingly, this bottom-up versus top-down connectivity profile of each of the layers is to some degree paralleled in functional data. That is, from neurophysiological and neuroimaging work, it is known that various visual stimuli and tasks can exert differential effects on the various layers \citep{Maier2010,Xing2012,Self2013,VelezFort2014, OHerron2016}. Intracranial work in monkeys, for instance, shows that for selective attention and working memory (two functions that are commonly associated with top-down processes), current source density is increased in deep and superficial compared to middle layers in primary visual cortex \citep{VanKerkoerle2017}. Similar layer specific patterns have been shown in animal functional MRI. For instance, whisker stimulation led to an increase in BOLD response in Layer IV of rat barrel cortex, before such an enhancement was observed in any of the other layers, suggesting that layer IV was the first to receive feed forward drive from upstream areas \citep{Yu2014}. In contrast, subsequent corticocortical connections in the same task appeared to activate Layers II-III and V in the motor cortex and contralateral barrel cortex before this affected any of the other layers, suggesting that these layers were the first to receive feedback signals. To what extent these results generalize to human cortex, however, remains to be investigated.

Recent advancements in fMRI have made it possible to also investigate the functional role of cortical layers in humans (e.g. \citet{Polimeni2010,Maass2014,Kok2016,Lawrence2018,Sharoh2019}).

The human in vivo resolution with fMRI has increased to submillimetre voxel size. The thickness of the cerebral cortex varies between 1 and 4.5 millimetres \citep{Zilles1990,Fischl2000}, giving sufficient resolution to characterise activity across the individual layers. Layer specific is now often used to try and measure layer specific activation in for example the visual system \citep{Muckli2015,Kok2016,Lawrence2018,DeHollander2020}, the motor system \citep{Huber2018}, working memory tasks \citep{Finn2019}, and to find directional connectivity between language areas \citep{Sharoh2019}. If layer specific analysis can make good on its promise of reliably discerning layer specific signals, it can be useful for answering questions in a wide range of cognitive domains \citep{Lawrence2019} and for questions of directional connectivity and cognitive network neuroscience \citep{Huber2020}.

@@

To facilitate reproducibility of our results, we further include a reproducible and openly accessible processing pipeline for layer-specific analyses. It includes benchmark tests for interactive visualisation of high resolution coregistration, cortical lamination, and cardiac and respiratory noise filtering. It is thereby a reproducible suggestion that may help optimize future study designs. Although we believe in the effectiveness of our pipeline, we encourage researchers to continue developing tools to improve the analysis with their preferred toolbox. We do encourage more consistency in the abstract steps of the analysis that are performed. It is therefore that we present our data analysis pipeline as an online visual workflow that can be easily inspected, shared, and adapted to fit any laminar study’s needs (\url{https://doi.org/10.5281/zenodo.3428603} \citep{VanMourik2018}). As a result of the relative novelty of fMRI investigations into individual cortical layers, previous work has used a large variety of high-resolution toolboxes and analyis pathways (LAYNII in \citet{Huber2017,Huber2018}; OpenFmriAnalysis in \citet{Lawrence2018}; Nighres in \citet{Huntenburg2018}; BrainVoyager in \citet{Goebel2012}), which has made direct comparison between studies difficult.

@@

We first focused on the effects of attention per se. Attention significantly enhanced the BOLD response at the attended location in areas V1-V3 (effect of attention, $F(1,16)=121.3$, $p=7.07 \cdot 10^{-9}$). The mean effect sizes (in percent signal change) were 1.01\%, 1.09\% and 0.96\% for V1, V2, and V3 respectively, and slightly stronger to those reported before (e.g., \citet{Murray2008,Jehee2011,Sprague2013}). To account for potential variation in baseline response between visual areas, and facilitate direct comparison with previous studies, we computed an Attentional Modulation Index (AMI; \citep{Kastner1999}. The AMI is defined as attentional effects normalised by the summed activity of both attended and unattended conditions. We found AMIs ($\mu \pm \sigma$) of 0.18 $\pm$ 0.05 for V1, 0.24 $\pm$ 0.07 for V2, and 0.24 $\pm$ 0.09 for V3. Direct comparison between areas revealed that although the absolute contribution of attention did not change, the relative contribution of attention was significantly higher in V2 and V3 than in V1 (AMI: F(2,32)=10.53, $p=3.07 \cdot 10^{-4}$, V1 compared to V2: $T(16)=-4.69$, $p=2.44 \cdot 10^{-3}$, V1 compared to V3 $T(16)=-3.17$, $p=5.89 \cdot 10^{-3}$, V2 compared to V3: $T(16)=-0.14$, $p=0.89$). Altogether, these results are  in line with previously reported effects of attention on coarse-level BOLD activity in visual cortex \citep{Somers1999,Gandhi1999}, and show that the cortical response for a spatial location is enhanced when attention is directed to that location.

@@

To  qualitatively assess the shape of the BOLD response over time and confirm the parameters of our GLM approach, we additionally conducted a Finite Impulse Response (FIR) analysis. The FIR analysis can be inspected and reproduced online in a Jupyter Notebook (\url{https://github.com/TimVanMourik/LayerAttention/blob/master/Notebooks/LayerFir.ipynb}). We extracted BOLD response curves for each experimental condition (see~\FIG{fir_global}), and observed clear and reliable effects of attention on the BOLD response that were fully consistent with our previous analyses. Because of the left-right modulation of attention, attentional effects were also reversed in the left and right hemisphere, as observed before and further illustrated in \FIGSUPP{fir_both_hemis}. Moreover, the cortical response over time for each condition was very well described by the canonical Hemodynamic Response Function (HRF) ($r^2=0.982$ with attention and $r^2=0.985$ without attention). Thus, the HRF model presented a fair description of the cortical responses observed in our experiment.

\begin{figure}
\includegraphics[width=\linewidth]{img/fir.png}
\caption{ The fitted BOLD response for each experimental condition. The shaded area represents the standard error of the mean over subjects. Results were obtained by fitting a Finite Impulse Response function of 18 time points to the BOLD data, starting at 2 seconds before and running until 15 seconds after stimulus onset. The dashed line indicates an HRF that was fitted to the FIR results of a pilot session. The obtained parameter values were used in the modelled HRF across layers in our GLM analyses (see Methods). These results are collapsed over hemisphere.}
\label{fig:fir_global}
%% If the optional argument in the square brackets is "none", then the caption *will not appear in the main figure at all* and only the full caption will appear under the supplementary figure at the end of the manuscript.
\figsupp[The results for the left and right hemisphere separately are displayed in the supplementary figures]{The fitted BOLD response for each experimental condition, separated by hemisphere. The shaded area represents the standard error of the mean over subjects. Results were obtained by fitting a Finite Impulse Response function of 18 time points, starting at 2 seconds before and running until 15 seconds after stimulus onset. The dashed line indicates an HRF that was fitted to the responses in a pilot session (see Methods). The same HRF parameter values were used in other statistical analysis.}{
% \includegraphics[width=\linewidth]{img/suppmat/fir_both_hemis.png}
}
\label{figsupp:fir_both_hemis}
\end{figure}

@@

To assess the significance of these effects, data were analyzed using a temporal general linear model with attention, stimulus, area, and layer as factors (see Methods). We found a reliable increase in BOLD response from white matter to pial surface (see \FIG{layer_region_stats_600}, overall effect of cortical depth, $F(2, 32)=87.5$, $p=1.07 \cdot 10^{-13}$). This increase in BOLD response with decreasing distance to the pial surface was reliably larger in the presence of a stimulus (two-way interaction between layer and stimulus, $F(2, 32)=85.6$, $p=1.43 \cdot 10^{-13}$)). Attention also led to reliable increases in BOLD response with decreasing distance to the pial surface (two-way interaction between layer and attention, $F(2,32)=43.10$, $p=8.34 \cdot 10{-10}$). 
This increase in BOLD response from lower to higher layers has been observed (e.g. \citet{Koopmans2010,Polimeni2010,Koopmans2011,Olman2012,Huber2018}) and modeled \cite{Markuerkiaga2016,Havlicek2020} previously, and is consistent with the blood flow from the gray-white matter boundary to the pial surface. That is, any change in BOLD response that arises in the deep layers will automatically affect responses in downstream layers, simply because blood flows from deeper to more superficial layers. The accumulation of blood in downstream layers can result in a larger slope of BOLD activation across the layers, even when there is no change in neural activity in these layers.
\begin{figure}
\includegraphics[width=\linewidth]{img/fir_layers.png}
\caption{The fitted BOLD response for each experimental condition for three cortical layers. The shaded area represents the standard error of the mean over subjects. Data were well described by the canonical HRF (dashed line).}
\label{fig:fir_layers}
\end{figure}

Next, we determined whether the layer-specific increase in BOLD signal was different for attention-based effects compared to those of the visual stimulus. If so, then this would be consistent with a targeted effect of attention on one of the layers. Interestingly, the attention-based increase in activity was reliably different from stimulus-driven changes in layer response (post hoc comparison between layer by stimulus effect and layer by attention effect; $T(16)=9.28$, p=$7.64\cdot10^{-8}$). This could potentially reflect a specific effect of attention on one of the layers. However, there is also an alternative interpretation, as the magnitude of an interaction effect is tightly coupled to the strength of its main effects in layer-specific analyses. 
This is because the interaction effect can be interpreted as a difference in slope (signal by depths) between two lines. Due to cortical signal leakage as a result of physiological (blood flow) and methodological reasons (errors in depth measurement), a higher visual stimulus response than an attentional response will be visible in \emph{both} the main effect \emph{and} the cortical slope of the response (see also above). Any reliable difference in slope is picked up as a significant interaction between layer by stimulus effect and layer by attention effect. However, this difference in slope would also arise if there was no layer-specific change in neural activity, so this finding alone does not indicate conclusive layer specific activation. Such conclusions might be made if a reliable three-way difference between the effects of attention and stimulus between any of the three layers would be present. However, we did not find a significant three-way interaction (three-way interaction between layer, stimulus and attention, $F(2,32)=0.96$, $p=0.393$), making it difficult to draw any firm conclusions regarding the layer-specific effects of the stimulus versus attention.

While we observed no clear and unambiguous laminar differences across visual areas, could there be a change within a given area? When analyzing stimulus activity, the pattern of activity across the layers was, indeed, significantly different between the three areas (three-way interaction between layer, stimulus, area: F(4, 64)=3.10, p=0.021)). Post hoc analyses revealed that the stimulus response across layers was slightly steeper in V1 than V2 (stimulus by layer effect in area V1 compared to V2, $T(16)=2.26$, $p=0.038$), while no significant difference was observed for area V1 compared to area V3 ($T(16)=1.35$, $p=0.196$), or V2 compared to V3 ($T(16)=-1.43$, $p=0.172$). This might reflect a slightly higher top layer activation in V1 compared to V2. On the other hand, it could also reflect the stronger main effect of stimulus in V1, leading to an increased slope of activation over layers, in line with expectations based on blood flow. We then focused on the layer-specific effects of attention in each individual area.
None of the equivalent analyses showed significant effects (attention by layer effect in area V1 compared to V2: $T(16)=0.14$, $p=0.89$, V1 compared to area V3: $T(16)=-0.88$, $p=0.39$, and V2 compared to V3: $T(16)=-1.43$, $p=0.17$). Thus, no region-specific attentional layer responses could be established.

Thus, while the overall effects on BOLD activity of both visual stimuli and attention were robust and similar to previously reported values for visual cortex (e.g. \citep{Kastner1999,Jehee2011, Koopmans2010}), both across and within layers, it proved more difficult to interpret the layer-specific pattern of activity for the two conditions. Although at first sight, it may appear that the stimulus affected the layers in a manner different from attention, the effect should be interpreted with caution because the stimulus also led to a stronger coarse-level cortical response than attention. 
We wanted to rule out that some important parameter choices obscured true effects. We verified robustness of our main results by changing the size of the region of interest and by varying the most important parameters of our layer extraction technique. We reprocessed the data and recomputed our layer and region specific statistical analysis for the experimental conditions: the ANOVA table of layer, region, stimulus, and attention. The control analyses established that the results were not strongly affected by the number of voxels included in the analyses (\FIGSUPP{layerresults300} and \FIGSUPP{layerresults900}), nor by the number of layers analyzed (\FIGSUPP{layerresults4layers}). In addition, the results did not qualitatively change when layer activation profiles were defined using volume interpolation (\FIGSUPP{layerresultsinterp}) as opposed to using a laminar spatial GLM as we did in the main analyses (see Methods). The control analyses and their comparison with the main analysis is further described in the supplemental materials.

@@

%% If the optional argument in the square brackets is "none", then the caption *will not appear in the main figure at all* and only the full caption will appear under the supplementary figure at the end of the manuscript.
\figsupp[none]{A control analysis that is identical to the main analysis, but repeated with a smaller region of interest. Only the 300 highest activated vertices were included in the analysis.}{\includegraphics[width=0.6\linewidth]{img/suppmat/layer_region_stats_300.png}}
\label{figsupp:layerresults300}
\figsupp[none]{A control analysis that is identical to the main analysis, but repeated with a larger region of interest. Only the 900 highest activated vertices were included in the analysis.}{\includegraphics[width=0.6\linewidth]{img/suppmat/layer_region_stats_900.png}}
\label{figsupp:layerresults900}
\figsupp[none]{A control analysis that is identical to the main analysis, but repeated with four extracted layers instead of three.}{\includegraphics[width=0.6\linewidth]{img/suppmat/layer_region_stats_4layers.png}}
\label{figsupp:layerresults4layers}
\figsupp[none]{A control analysis that is identical to the main analysis, but repeated with a different way of obtaining laminar signal.}{}{
\includegraphics[width=0.6\linewidth]{img/suppmat/layer_region_stats_interp.png}
}
\label{figsupp:layerresultsinterp}

@@

Why did we not find a targeted effect of attention on one of the layers? One possibility is that our data are simply insufficiently robust to show a significant difference in activity across depth between the two conditions. It is well known that the BOLD signal includes multiple sources of noise related to both MRI scanner and participant, and this holds especially true for signals recorded at the sub-millimeter scale. For example, at a resolution this high, even the smallest movement of the participant may cause additional blurring of the data, with potentially detrimental effects on the signal-to-noise ratio. For this reason, we collected data from 17 participants - a sample size much larger than typical in attention-based fMRI studies at standard spatial resolution (cf. N=4-6 in \citet{Kastner1999,Kamitani2005,Jehee2011}), and on the high end compared to layer-based fMRI studies at high resolution (cf. N=6 in \citet{Polimeni2010}, N=4 in \citet{Muckli2015}, N=10 in \citet{Kok2016}, N=12 in \citep{Klein2018}, N=21 in \citep{Lawrence2018}, N=22 in \citep{Sharoh2019}). To minimize the effects of various sources of noise, we took great care in measuring and removing physiological artifacts, and further improved existing layer extraction techniques by developing a novel technique to separate laminar signal from different layers \cite{VanMourik2019}. Indeed, the combined success of these procedures is well illustrated by the effect sizes observed in the current study for both stimulus presentation (4.5\%, 3.3\%, 2.8\% in V1, V2 and V3) and attention (0.41\%, 0.64\%, 0.59\% in V1, V2 and V3), which are comparable or higher to those reported in previous work \citet{Murray2008,Jehee2011}. We also ensured that similar results were obtained using more conventional layer-extraction procedures, and that the results were robust to changes in signal extraction method or number of layers. 
There are, however, some differences in experimental design between our study and previous laminar investigations that could potentially account for the incongruity in results. Because we were interested in the degree to which top-down processes could be dissociated from feed forward stimulation with fMRI, we directly contrasted between these two conditions in our analyses. Previous studies, on the other hand, have focused on top-down activity in isolation (e.g. \citet{Muckli2015}, \citet{Kok2016}, \citet{Lawrence2018}). Two-by-two experimental designs are surprisingly rare in layer-specific analysis and have only been recently employed by \citet{DeHollander2020}. We emphasise that a multifactorial design is important to account for all the degrees of freedom as introduced by activation, cortical depth, signal leakage, and analytical flexibility. A more limited experimental design might force an analysis into single-layer comparisons, which might lead to oversimplifications and incorrect conclusions about differential laminar activation. 
Alternate strategies are to compare laminar signal for information content between conditions by means of multi-voxel pattern analysis \citet{Muckli2015}; for retinotopic preference in the visual cortex \citep{Klein2018}; or by means of a general psycho-physiological interaction \citep{Sharoh2019} to shed light on inter-regional laminar communication. It will be interesting for future studies to directly compare between these and other choices in analysis approach to see which matter most for robust, replicable, and meaningful laminar results.

>>>>

\contrib[\authfn{1}]{These authors contributed equally to this work}
\contrib[\authfn{2}]{These authors contributed equally to this work}

@@

\begin{abstract}
Directing spatial attention towards a particular stimulus enhances cortical responses at corresponding regions in cortex. How attention modulates the laminar response profile within the attended region, however, remains unclear. In this paper, we use high field (7T) fMRI to investigate the effects of attention on laminar activity profiles in areas V1-V3; both when a stimulus was presented to the observer, and in the absence of visual stimulation. Replicating previous findings, we find robust increases in the overall BOLD response for attended regions in cortex, both with and without visual stimulation. When analyzing the BOLD response across the individual layers in visual cortex, we observed no evidence for laminar-specific differentiation with attention. We offer several potential explanations for these results, including theoretical, methodological and technical reasons. Additionally, we provide all data and pipelines openly, in order to promote analytic consistency across layer-specific studies, improve reproducibility, and decrease the false positive rate as a result of analytical flexibility.
\end{abstract}

@@

It is known from anatomical studies that the human cerebral cortex can be subdivided into histological layers with different cell types. The cytoarchitectonic structure varies across the brain and forms the basis of the Brodmann atlas \citep{Brodmann1909}. While the precise function of each cortical layer remains unclear, their connectivity profile suggests a division in terms of bottom-up and top-down processing \citep{Felleman1991,Barone2000,Shipp2016}. Most brain areas have six different histological layers. Specifically, Layer IV and to a lesser extent Layer V/VI are commonly associated with receiving feedforward drive from Layer III of lower cortical areas or from the thalamus \citep{Jones1998,Constantinople2013}. Layers I-II and VI, in contrast, are typically implicated in receiving downward information flow (feedback), which often originates from layer V \citep{Alitto2003}. This bottom-up versus top-down connectivity profile of each of the layers is to some degree also paralleled in functional data. That is, from neurophysiological and neuroimaging work, it is known that various visual stimuli and tasks can exert differential effects on the various layers \citep{Maier2010,Xing2012,Self2013,VelezFort2014, OHerron2016}. Intracranial work in monkeys, for instance, shows that for selective attention and working memory (two functions that are commonly associated with top-down processes), current source density is increased in deep and superficial compared to middle layers in primary visual cortex \citep{VanKerkoerle2017}. Similar layer specific patterns have been shown in animal functional MRI. For instance, whisker stimulation led to an increase in BOLD response in Layer IV of rat barrel cortex, before such an enhancement was observed in any of the other layers, suggesting that layer IV was the first to receive feed forward drive from upstream areas \citep{Yu2014}. In contrast, subsequent cortico-cortical connections in the same task appeared to activate Layers II-III and V in the motor cortex and contralateral barrel cortex before this affected any of the other layers, suggesting that these layers were the first to receive feedback signals. To what extent these results generalize to human cortex, however, remains to be investigated.

Recent advancements in fMRI have made it possible to also investigate the functional role of cortical layers in humans (e.g. \citet{Polimeni2010,Maass2014,Kok2016,Lawrence2018,Sharoh2019}). The human in vivo resolution with fMRI has increased to submillimetre voxel size. The thickness of the cerebral cortex varies between 1 and 4.5 millimetres \citep{Zilles1990,Fischl2000}, giving sufficient resolution to characterise activity across the individual layers. FMRI is now often used to try and measure layer specific activation in, for example, the visual system \citep{Muckli2015,Kok2016,Lawrence2018,DeHollander2020}, the motor system \citep{Huber2018}, working memory tasks \citep{Finn2019}, and to find directional connectivity between language areas \citep{Sharoh2019}. If layer specific analysis can make good on its promise of reliably discerning layer specific signals, it can be useful for answering questions in a wide range of cognitive domains \citep{Lawrence2019} and for questions of directional connectivity and cognitive network neuroscience \citep{Huber2020}, including research questions involving spatial attention.

@@

To facilitate reproducibility of our results, we further include a reproducible and openly accessible processing pipeline for layer-specific analyses (\url{https://doi.org/10.5281/zenodo.3428603} \citep{VanMourik2018}). The toolbox includes benchmark tests for interactive visualisation of high resolution coregistration, cortical lamination, and cardiac and respiratory noise filtering. To enhance transparency, the data analysis pipeline is furthermore presented as an online visual workflow that can be easily inspected, shared, and adapted to fit any laminar study’s needs. As a result of the relative novelty of fMRI investigations into individual cortical layers, previous work has used a large variety of high-resolution toolboxes and analyis pathways (LAYNII in \citet{Huber2017,Huber2018}; OpenFmriAnalysis in \citet{Lawrence2018}; Nighres in \citet{Huntenburg2018}; BrainVoyager in \citet{Goebel2012}), which has made direct comparison between studies difficult. We hope our analysis pipeline will help ameliorate this issue, although we readily recognize that other toolboxes and solutions could similarly achieve such goals.

@@

We first focused on the effects of attention per se. Attention significantly enhanced the BOLD response at the attended location in areas V1-V3 (effect of attention, $F(1,16)=121.3$, $p=7.07 \cdot 10^{-9}$). The mean effect sizes (in percent signal change) were 1.01\%, 1.09\% and 0.96\% for V1, V2, and V3 respectively, and slightly stronger to those reported before (e.g., \citet{Murray2008,Jehee2011,Sprague2013}). To account for potential variation in baseline response between visual areas, and facilitate direct comparison with previous studies, we computed an Attentional Modulation Index (AMI; \citep{Kastner1999}. The AMI is defined as attentional effects normalised by the summed activity of both attended and unattended conditions. We found AMIs ($\mu \pm \sigma$) of 0.18 $\pm$ 0.05 for V1, 0.24 $\pm$ 0.07 for V2, and 0.24 $\pm$ 0.09 for V3. Direct comparison between areas revealed that although the absolute contribution of attention did not change, the relative contribution of attention differed significantly between regions (AMI: F(2,32)=10.53, $p=3.07 \cdot 10^{-4}$. Post-hoc comparison showed significant differences in V1 compared to V2: $T(16)=-4.69$, $p=2.44 \cdot 10^{-3}$ and in V1 compared to V3 $T(16)=-3.17$, $p=5.89 \cdot 10^{-3}$, but not in V2 compared to V3: $T(16)=-0.14$, $p=0.89$). Altogether, these results are  in line with previously reported effects of attention on coarse-level BOLD activity in visual cortex \citep{Somers1999,Gandhi1999}, and show that the cortical response for a spatial location is enhanced when attention is directed to that location.

@@

To  qualitatively assess the shape of the BOLD response over time and confirm the parameters of our GLM approach, we additionally conducted a Finite Impulse Response (FIR) analysis. The FIR analysis can be inspected and reproduced online in a Jupyter Notebook (\url{https://github.com/TimVanMourik/LayerAttention/blob/master/Notebooks/LayerFir.ipynb}). We extracted BOLD response curves for each experimental condition (see~\FIG{fir_global}), and observed clear and reliable effects of attention on the BOLD response that were fully consistent with our previous analyses. Because of the left-right modulation of attention, attentional effects were reversed in the left and right hemisphere, as observed before and further illustrated in \FIGSUPP{fir_both_hemis}. Moreover, the cortical response over time for each condition was very well described by the canonical Hemodynamic Response Function (HRF) ($r^2=0.982$ with attention and $r^2=0.985$ without attention). Thus, the HRF model presented a fair description of the cortical responses observed in our experiment.

\begin{figure}
\includegraphics[width=\linewidth]{img/fir.png}
\caption{ The fitted BOLD response for each experimental condition. The shaded area represents the standard error of the mean over subjects. Results were obtained by fitting a Finite Impulse Response function of 18 time points to the BOLD data, starting at 2 seconds before and running until 15 seconds after stimulus onset. The dashed line indicates an HRF that was fitted to the FIR results of a pilot session. The obtained parameter values were used in the modelled HRF across layers in our GLM analyses (see Methods). These results are collapsed over hemisphere.}
\label{fig:fir_global}
%% If the optional argument in the square brackets is "none", then the caption *will not appear in the main figure at all* and only the full caption will appear under the supplementary figure at the end of the manuscript.
\figsupp[The results for the left and right hemisphere separately are displayed in the supplementary figures]{The fitted BOLD response for each experimental condition, separated by hemisphere. The shaded area represents the standard error of the mean over subjects. Results were obtained by fitting a Finite Impulse Response function of 18 time points, starting at 2 seconds before and running until 15 seconds after stimulus onset. The dashed line indicates an HRF that was fitted to the responses in a pilot session (see Methods). The same HRF parameter values were used in other statistical analysis.}{\includegraphics[width=\linewidth]{img/suppmat/fir_both_hemis.png}}
\label{figsupp:fir_both_hemis}
\end{figure}

@@

To assess the significance of these effects, data were analyzed using a temporal general linear model with attention, stimulus, area, and layer as factors (see Methods). We found a reliable increase in BOLD response from white matter to pial surface (see \FIG{layer_region_stats_600}, overall effect of cortical depth, $F(2, 32)=87.5$, $p=1.07 \cdot 10^{-13}$). This increase in BOLD response with decreasing distance to the pial surface was reliably larger in the presence of a stimulus (two-way interaction between layer and stimulus, $F(2, 32)=85.6$, $p=1.43 \cdot 10^{-13}$)). Attention also led to reliable increases in BOLD response with decreasing distance to the pial surface (two-way interaction between layer and attention, $F(2,32)=43.10$, $p=8.34 \cdot 10{-10}$). 
This increase in BOLD response from lower to higher layers has been observed (e.g. \citet{Koopmans2010,Polimeni2010,Koopmans2011,Olman2012,Huber2018}) and modeled \citep{Markuerkiaga2016,Havlicek2020} previously, and is consistent with the blood flow from the gray-white matter boundary to the pial surface. That is, any change in BOLD response that arises in the deep layers will automatically affect responses in downstream layers, simply because blood flows from deeper to more superficial layers. The accumulation of blood in downstream layers can result in a larger slope of BOLD activation across the layers, even when there is no change in neural activity in these layers.
\begin{figure}
\includegraphics[width=\linewidth]{img/fir_layers.png}
\caption{The fitted BOLD response for each experimental condition for three cortical layers. The shaded area represents the standard error of the mean over subjects. The horizontal axis represents time (in seconds) from stimulus onset.}
\label{fig:fir_layers}
\end{figure}

Next, we determined whether the layer-specific increase in BOLD signal was different for attention-based effects compared to those of the visual stimulus. If so, then this would be consistent with a targeted effect of attention on one of the layers. Interestingly, the attention-based increase in activity was reliably different from stimulus-driven changes in layer response (post hoc comparison between layer by stimulus effect and layer by attention effect; $T(16)=9.28$, p=$7.64\cdot10^{-8}$). This could potentially reflect a specific effect of attention on one of the layers. However, there is also an alternative interpretation, as the magnitude of an interaction effect is tightly coupled to the strength of its main effects in layer-specific analyses. 
This is because the interaction effect can be interpreted as a difference in slope (signal by depths) between two lines. Due to cortical signal leakage as a result of physiological (blood flow) and methodological reasons (errors in depth measurement), a higher visual stimulus response than an attentional response will be visible in \emph{both} the main effect \emph{and} the cortical slope of the response (see also above). Any reliable difference in slope is picked up as a significant difference between layer by stimulus effect and layer by attention effect. However, this difference in slope would also arise if there was no layer-specific change in neural activity, so this finding alone does not indicate conclusive layer specific activation. Such conclusions might be made if a reliable three-way difference between the effects of attention and stimulus between any of the three layers would be present. However, we did not find a significant three-way interaction (three-way interaction between layer, stimulus and attention, $F(2,32)=0.96$, $p=0.393$), making it difficult to draw any firm conclusions regarding the layer-specific effects of the stimulus versus attention.

While we observed no clear and unambiguous laminar differences across visual areas, could there be a change within a given area? When analyzing stimulus activity, the pattern of activity across the layers was, indeed, significantly different between the three areas (three-way interaction between layer, stimulus, area: F(4, 64)=3.10, p=0.021)). Post hoc analyses revealed that the stimulus response across layers was slightly steeper in V1 than V2 (stimulus by layer effect in area V1 compared to V2, $T(16)=2.26$, $p=0.038$), while no significant difference was observed for area V1 compared to area V3 ($T(16)=1.35$, $p=0.196$), or V2 compared to V3 ($T(16)=-1.43$, $p=0.172$). This might reflect a slightly higher top layer activation in V1 compared to V2. On the other hand, it could also reflect the stronger main effect of stimulus in V1, leading to an increased slope of activation over layers, in line with expectations based on blood flow. We then focused on the layer-specific effects of attention in each individual area.
Attention does not vary layer specifically per region (attention by layer by region effect $F(64,4)=0.998$, $p=0.416$).

Thus, while the overall effects on BOLD activity of both visual stimuli and attention were robust and similar to previously reported values for visual cortex (e.g. \citep{Kastner1999,Jehee2011, Koopmans2010}), both across and within layers, it proved more difficult to interpret the layer-specific pattern of activity for the two conditions. Although at first sight, it may appear that the stimulus affected the layers in a manner different from attention (when analyzed across areas), the effect should be interpreted with caution because the stimulus also led to a stronger coarse-level cortical response than attention. 
We wanted to rule out that some important parameter choices obscured true effects. We verified robustness of our main results by changing the size of the region of interest and by varying the most important parameters of our layer extraction technique. We reprocessed the data and recomputed our layer and region specific statistical analysis for the experimental conditions. The control analyses established that the results were not strongly affected by the number of voxels included in the analyses (\FIGSUPP{layerresults300} and \FIGSUPP{layerresults900}), nor by the number of layers analyzed (\FIGSUPP{layerresults4layers}). In addition, the results did not qualitatively change when layer activation profiles were defined using volume interpolation (\FIGSUPP{layerresultsinterp}) as opposed to using a laminar spatial GLM as we did in the main analyses (see Methods). The control analyses and their comparison with the main analysis is further described in the supplemental materials.

@@

%% If the optional argument in the square brackets is "none", then the caption *will not appear in the main figure at all* and only the full caption will appear under the supplementary figure at the end of the manuscript.
\figsupp[none]{A control analysis that is identical to the main analysis, but repeated with a smaller region of interest. Only the 300 highest activated vertices were included in the analysis.}{\includegraphics[width=0.6\linewidth]{img/suppmat/layer_region_stats_300.png}}
\label{figsupp:layerresults300}
\figsupp[none]{A control analysis that is identical to the main analysis, but repeated with a larger region of interest. Only the 900 highest activated vertices were included in the analysis.}{\includegraphics[width=0.6\linewidth]{img/suppmat/layer_region_stats_900.png}}
\label{figsupp:layerresults900}
\figsupp[none]{A control analysis that is identical to the main analysis, but repeated with four extracted layers instead of three.}{\includegraphics[width=0.6\linewidth]{img/suppmat/layer_region_stats_4layers.png}}
\label{figsupp:layerresults4layers}
\figsupp[none]{A control analysis that is identical to the main analysis, but repeated with a different way of obtaining laminar signal.}{\includegraphics[width=0.6\linewidth]{img/suppmat/layer_region_stats_interp.png}}
\label{figsupp:layerresultsinterp}

@@

Why did we not find a targeted effect of attention on one of the layers? One possibility is that our data are simply insufficiently robust to show a significant difference in activity across depth between the two conditions. It is well known that the BOLD signal includes multiple sources of noise related to both MRI scanner and participant, and this holds especially true for signals recorded at the sub-millimeter scale. For example, at a resolution this high, even the smallest movement of the participant may cause additional blurring of the data, with potentially detrimental effects on the signal-to-noise ratio. For this reason, we collected data from 17 participants - a sample size much larger than typical in attention-based fMRI studies at standard spatial resolution (cf. N=4-6 in \citet{Kastner1999,Kamitani2005,Jehee2011}), and on the high end compared to layer-based fMRI studies at high resolution (cf. N=6 in \citet{Polimeni2010}, N=4 in \citet{Muckli2015}, N=10 in \citet{Kok2016}, N=12 in \citep{Klein2018}, N=21 in \citep{Lawrence2018}, N=22 in \citep{Sharoh2019}). To minimize the effects of various sources of noise, we took great care in measuring and removing physiological artifacts, and further improved existing layer extraction techniques by developing a novel technique to separate laminar signal from different layers \citep{VanMourik2019}. Indeed, the combined success of these procedures is well illustrated by the effect sizes observed in the current study for both stimulus presentation (4.5\%, 3.3\%, 2.8\% in V1, V2 and V3) and attention (0.41\%, 0.64\%, 0.59\% in V1, V2 and V3), which are comparable or higher to those reported in previous work \citep{Murray2008,Jehee2011}. We also ensured that similar results were obtained using more conventional layer-extraction procedures, and that the results were robust to changes in signal extraction method or number of layers. 
There are, however, some differences in experimental design between our study and previous laminar investigations that could potentially account for the incongruity in results. Because we were interested in the degree to which top-down processes could be dissociated from feed forward stimulation with fMRI, we directly contrasted between these two conditions in our analyses. Previous studies, on the other hand, have focused on top-down activity in isolation (e.g. \citet{Muckli2015}, \citet{Kok2016}, \citet{Lawrence2018}). Two-by-two experimental designs are surprisingly rare in layer-specific analysis and have only been recently employed by \citet{DeHollander2020}. We emphasise that a multifactorial design is important to account for changes due to, for example, cortical depth and signal leakage per se, as opposed to true layer-based changes in activity due to the experimental manipulations. Alternate strategies are to compare between layers and conditions in terms of information content \citep{Muckli2015}, retinotopic preference \citep{Klein2018}, or by focusing on inter-regional laminar communication \citep{Sharoh2019}. For example, \citet{Klein2018} recently observed a cortical depth-dependent shift in population receptive fields with spatial attention. This raises the intriguing possibility that, in our study, spatial attention led to a depth-dependent shift in the strength of orientation-selective responses – indeed, previous coarse-scale fMRI studies have observed that orientation selectivity can change even when there is no change in amplitude across the population \citep{Jehee2011,Jehee2012}.
